An Intrinsically Typed Solution for the List Machine Benchmark

### 2020-06-23 papers/SBLP2020/paper.tex
Starting the related work

A popular approach for specifying formal semantics is the
use of definitional interpreters, which represents the meaning of a
programming language as an interpreter written in some
meta-language~\cite{Reynolds72}. A major advantage of such approach
is the possibility of validating the semantics through execution.
Recently, definitional interpreters were used to formalize type
soundness theorems for some advanced typing features~\cite{Amin17}
and the semantics of imperative programming languages in which static
semantics is ensured by dependently-typed
syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
In this work, we follow Poulsen et. al and use an intrinsically-typed
representation to build a definitional interpreter for a low-level virtual
machine developed by Appel and Leroy as a benchmark problem closer to
real-world implementations like typed assembly languages~\cite{CraryM99} and
proof carrying code~\cite{Necula97}.

@@

A key feature of the list machine type system is its subypting which is exposed by the
the least common supertype relation. We omit the Agda encoding of the subtyping relation for
space reasons, since it is just an inductive type that encode the rules presented
in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
subtyping including its decidability.

@@

\paragraph{Definitional intepreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
}

>>>>

A popular approach for specifying formal semantics is the
use of definitional interpreters, which represents the meaning of a
programming language as an interpreter written in some
meta-language~\cite{Reynolds72}. A major advantage of such approach
is the possibility of validating the semantics through execution.
Recently, definitional interpreters were used to formalize type
soundness theorems for some advanced typing features~\cite{Amin17}
and the semantics of imperative programming languages in which static
semantics is ensured by dependently-typed
syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
In this work, we follow Poulsen et. al. and use an intrinsically-typed
representation to build a definitional interpreter for a low-level virtual
machine developed by Appel and Leroy as a benchmark problem closer to
real-world implementations like typed assembly languages~\cite{CraryM99} and
proof carrying code~\cite{Necula97}.

@@

A key feature of the list machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
space reasons, since it is just an inductive type that encode the rules presented
in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
subtyping including its decidable test.

@@

\paragraph{Definitional intepreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
Recently, the interest on such interpreters was revitalized by Amin et. al.~\cite{Amin17} which used definitional
interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
Unlike our work, Amin et. al. formalizations does not use intrinsically-typed representations of their syntax which cluttered
their formalization with the need of dealing with ``stuck states'' in the semantics. Poulsen el. al. described how
definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
using a library for name binding using \emph{scope graphs}~\cite{Poulsen18}, which greatly simplifies the treatment of complex
binding structures. Since, the list machine benchmark was designed to address other problems than binding, our representation
using \emph{de Bruijn} indexes was sufficient to implement the desired typechecker and interpreter.
Other recent application of definitional interpreters was proposed by Rouvoet et. al.~\cite{Rouvoet20}. The main contribution
of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
linear typing features in the intepreter, authors have implemented monads based on a separation algebra which they
named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
Another work using intrisincally-typed syntax for resource control was developed by Thiemann~\cite{Thiemann19}, which
implemented an interpreter for a more realistic core functional session typed calculus including recursion and
session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
simple interface to a scheduler.
}
